Place of death in NRS COVID-19 deaths: data processing for analysis, 2023 data

### 2023-02-13 covid_nrs_place_of_death_data_2023.Rmd
moved section for historic cause of death data around for rclarity; finished processing historic period data for different years

```{r}
# TODO: deal with 2022 data here, now that it's 'historical'
cell_ranges_for_causes_of_death_data_2022 <- list(
  "All" = "A6:V58",
  "Care home" = "A61:V113",
  "Home & other non-institution" = "A116:V168",
  "Hospital" = "A171:V223",
  "Other" = "A226:V278"
)

@@

#### Historical causes of death

There is a second sense of 'historical' here, which denotes cause of death figures for 2020 to 2022; these were only provided as 5-year averages for 2015-2019, whereas we have a per-cause breakdown from 2020 onwards - to combine these, a 5-year historical average needs to be recomputed using a combination of the 2015-2019 data and the 2020+ data (using weighted averages).

Because the data from July 2022 onwards are split up so that I load the 2020 and 2021 from different sources, I'll compute the **Note: I don't remember what came after this...**

```{r}
## this is an ugly hack - we set average weekly deaths due to C19 to NA for 2021
## and we set the average covid deaths for 2015-2019, 2021 to the same as in 2021
## however, for the latter, we set the dates to 2022
# average_weekly_covid_deaths_2021_onwards <-
average_weekly_covid_deaths_2021_onwards <-
  bind_rows(
    ## set 2021 averages to NA
    weekly_deaths_cause_of_death_2021 %>%
    filter(cause_of_death == "COVID-19") %>%
    mutate(number_of_deaths = NA_real_) %>%
    rename(deaths_mean = number_of_deaths),
    ## use 2021 weekly deaths as 2022 historic average
    weekly_deaths_cause_of_death_2021 %>%
    filter(cause_of_death == "COVID-19") %>%
    mutate(year = 2022) %>%
    select(-c(week_number_run_over, date_w_c)) %>%
    rename(deaths_mean = number_of_deaths) %>%
    inner_join(weekly_deaths_cause_of_death_2022_onwards %>% 
                select(year, week_number, week_number_run_over, date_w_c, place_of_death, cause_of_death)
              )
  ) %>%
  mutate(historic_period = historic_period_depending_on_year(year))

#### Merge 2020 & 2021 data with current data

@@

@@

week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks; averages for week 53 were computed from figures for week 53 in 2015 and week 52 for 2016-2019."

# TODO: weekly averages may be too detailed for this level od subcategorising!

deaths_mean_min_max <- function(data_tbl) {
  data_tbl %>%
    summarise(
      deaths_mean = mean(number_of_deaths),
      deaths_min = min(number_of_deaths),
      deaths_max = max(number_of_deaths),
      .groups = "drop"
    )
}
```

@@

```{r}
historical_weekly_average_hb <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_hb, 
        current_data = weekly_deaths_hb,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, ref_area
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates

@@

```{r}
# TODO: include the split between historical periods here also
compute_historical_annual_average_deaths <- function(data_tbl, ...) {
  grouping_vars <- enquos(...)
  data_tbl %>%
    group_by(year, place_of_death, historic_period, !!!grouping_vars) %>%
    summarise(  # calculate annual totals first...
    number_of_deaths = sum(number_of_deaths), 
    .groups = "keep"
  ) %>%
  ungroup(year) %>%  # remove year as grouping to average over year now
  summarise(
    annual_deaths_mean = mean(number_of_deaths),
    annual_deaths_min = min(number_of_deaths),
    annual_deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
}

historical_annual_average_overall <-
  bind_rows(
      ## 2016-2019
      weekly_deaths_2015_2019_overall %>%
        mutate(year = as.integer(year)) %>%
        filter(year %in% 2016:2019) %>%
        mutate(historic_period = historic_period_depending_on_year(2020)),
      ## 2021
      weekly_deaths_overall %>% 
        rename(number_of_deaths = deaths_all_causes) %>%
        filter(year == 2021 & age == "all" & sex == "all") %>%
        mutate(historic_period = historic_period_depending_on_year(2022))
      ) %>%
  compute_historical_annual_average_deaths()

make_2016_to_2019_and_2021_data_with_historic_periods <- function (historical_data, current_data) {
  bind_rows(
    historical_data %>% 
      filter(year %in% 2016:2019) %>%
      mutate(historic_period = historic_period_depending_on_year(2020)),
    current_data %>% 
      filter(year == 2021) %>% 
      rename(number_of_deaths = deaths_all_causes) %>%
      mutate(historic_period = historic_period_depending_on_year(2022))
  )
}

historical_annual_average_hb <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_hb, 
    current_data = weekly_deaths_hb
      ) %>%
  compute_historical_annual_average_deaths(ref_area)

historical_annual_average_la <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_la, 
    current_data = weekly_deaths_la
      ) %>%
  compute_historical_annual_average_deaths(ref_area)

historical_annual_average_sex_age <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_sex_age, 
    current_data = weekly_deaths_sex_age
      ) %>%
  compute_historical_annual_average_deaths(sex, age)

historical_annual_average_sex <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_sex, 
    current_data = weekly_deaths_sex
      ) %>%
  compute_historical_annual_average_deaths(sex)

historical_annual_average_age <-
  make_2016_to_2019_and_2021_data_with_historic_periods(
    historical_data = weekly_deaths_2015_2019_age, 
    current_data = weekly_deaths_age
      ) %>%
  compute_historical_annual_average_deaths(age)
```

@@

>>>>

```{r}
# TODO: deal with 2022 data here, now that it's 'historical'; this will involve loading the last released data for 2022 released in the weekly releases for 2023, I assume somewhere mid-2023, like it was for 2021 data in 2022 releases
cell_ranges_for_causes_of_death_data_2022 <- list(
  "All" = "A6:V58",
  "Care home" = "A61:V113",
  "Home & other non-institution" = "A116:V168",
  "Hospital" = "A171:V223",
  "Other" = "A226:V278"
)

@@

@@

# Historic data for deaths

There is a second sense of 'historical' here, which denotes cause of death figures for 2020 to 2022; these were only provided as 5-year averages for 2015-2019, whereas we have a per-cause breakdown from 2020 onwards - to combine these, a 5-year historical average needs to be recomputed using a combination of the 2015-2019 data and the 2020+ data (using weighted averages). This is done in [the later section recomputing historical averages for causes of death](#historical-annual-average-by-cause).

@@

week_53_text <- "Due to the way ISO 8601 counts weeks, only 2015 in the 2015-2019 period had 53 weeks; averages for week 53 were computed from figures for week 53 in 2015 and week 52 for 2016-2019."
```

@@

```{r}
## helper function to summarise with mean, min, and max
deaths_mean_min_max <- function(data_tbl) {
  data_tbl %>%
    summarise(
      deaths_mean = mean(number_of_deaths),
      deaths_min = min(number_of_deaths),
      deaths_max = max(number_of_deaths),
      .groups = "drop"
    )
}

historical_weekly_average_hb <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_hb, 
        current_data = weekly_deaths_hb,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, ref_area
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates

@@

Here I'm computing the 5-year annual average for various groupings of variables, and across the various possible historic periods. Instead of taking the weekly averages for a historic period computed in the above section, I recompute this separately from the raw data for historic periods and current data.

```{r}
## helper function to compute annual average per historic period, per place of
## death, and per various grouping variables as needed
compute_historical_annual_average_deaths <- function(data_tbl, ...) {
  grouping_vars <- enquos(...)
  data_tbl %>%
    group_by(year, place_of_death, historic_period, !!!grouping_vars) %>%
    summarise(  # calculate annual totals first...
    number_of_deaths = sum(number_of_deaths), 
    .groups = "keep"
  ) %>%
  ungroup(year) %>%  # remove year as grouping to average over year now
  summarise(
    annual_deaths_mean = mean(number_of_deaths),
    annual_deaths_min = min(number_of_deaths),
    annual_deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
}

## Note, there was an error in these before, where I erroneously omitted 2015
## figures from the annual averages computation for 2015-2019; this had minimal
## effect on the mean/min/max values because the algorithm used averaged over
## the available years (as opposed to e.g. dividing 4 years of data by 5)

# the below commented out section is no longer used, but I've left it to
# demonstrate the error in previous data - I think I started 'splitting' the
# 2015-2019 afrom the 2016-2021 historic periods but didn't end up implementing
# both, and instead left a confusing entry that compute the 2016-2019 period
# only.
# make_2016_to_2019_and_2021_data_with_historic_periods <- function(historical_data, current_data) {
#   bind_rows(
#     historical_data %>% 
#       filter(year %in% 2016:2019) %>%
#       mutate(historic_period = historic_period_depending_on_year(2020)),
#     current_data %>% 
#       filter(year == 2021) %>% 
#       rename(number_of_deaths = deaths_all_causes) %>%
#       mutate(historic_period = historic_period_depending_on_year(2022))
#   )
# }

historical_annual_average_overall <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = 
          weekly_deaths_2015_2019_overall %>%  # a quirk of the weekly deaths 'overall' data
          filter(year != "average") %>%
          mutate(year = as.integer(year)), 
        current_data = 
          weekly_deaths_overall %>%
          filter(sex == "all" & age == "all"),
        year_range = years_in_historic_period
      ) %>%
        mutate(historic_period = historic_period)
    }
  ) %>%
  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
  compute_historical_annual_average_deaths()

historical_annual_average_hb <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_hb, 
        current_data = weekly_deaths_hb, 
        year_range = years_in_historic_period
      ) %>%
        mutate(historic_period = historic_period)
    }
  ) %>%
  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
  compute_historical_annual_average_deaths(ref_area)

historical_annual_average_la <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_la, 
        current_data = weekly_deaths_la, 
        year_range = years_in_historic_period
      ) %>%
        mutate(historic_period = historic_period)
    }
  ) %>%
  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
  compute_historical_annual_average_deaths(ref_area)

historical_annual_average_sex_age <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_sex_age, 
        current_data = weekly_deaths_sex_age, 
        year_range = years_in_historic_period
      ) %>%
        mutate(historic_period = historic_period)
    }
  ) %>%
  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
  compute_historical_annual_average_deaths(sex, age)

historical_annual_average_sex <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_sex, 
        current_data = weekly_deaths_sex, 
        year_range = years_in_historic_period
      ) %>%
        mutate(historic_period = historic_period)
    }
  ) %>%
  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
  compute_historical_annual_average_deaths(sex)

historical_annual_average_age <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_age, 
        current_data = weekly_deaths_age, 
        year_range = years_in_historic_period
      ) %>%
        mutate(historic_period = historic_period)
    }
  ) %>%
  distinct %>%  # 2020 and 2021 use the same historic period, so I remove duplicate entries at this point
  compute_historical_annual_average_deaths(age)
```

# Historic average by cause of death

```{r}
# TODO: change this so it works for 2023 and onwards

## this is an ugly hack - we set average weekly deaths due to C19 to NA for 2021
## and we set the average covid deaths for 2015-2019, 2021 to the same as in 2021
## however, for the latter, we set the dates to 2022
average_weekly_covid_deaths_2021_onwards <-
  bind_rows(
    ## set 2021 averages to NA
    weekly_deaths_cause_of_death_2021 %>%
    filter(cause_of_death == "COVID-19") %>%
    mutate(number_of_deaths = NA_real_) %>%
    rename(deaths_mean = number_of_deaths),
    ## use 2021 weekly deaths as 2022 historic average
    weekly_deaths_cause_of_death_2021 %>%
    filter(cause_of_death == "COVID-19") %>%
    mutate(year = 2022) %>%
    select(-c(week_number_run_over, date_w_c)) %>%
    rename(deaths_mean = number_of_deaths) %>%
    inner_join(weekly_deaths_cause_of_death_2022_onwards %>% 
                select(year, week_number, week_number_run_over, date_w_c, place_of_death, cause_of_death)
              )
  ) %>%
  mutate(historic_period = historic_period_depending_on_year(year))

## Merge 2020 & 2021 data with current data

@@

Cause of death data for 2015-2019 was only given in terms of weekly averages over 5 years. With the later historic periods (2016-2019 + 2021 and so on) I compute a weighted average between the 5-year average and the current weekly data.

The below compromise figure is no longer used, but the reasoning for the weighted average is the same (except for the quirk of the denominator of 6 in the below case, where I'm including the 5 years of historic data plus the one year of current data, hence 6 years in the denominator):
